Think deep for reef resilience

### 2025-08-17 paper/paper_draft.qmd
Mainly correctly explain depth ranges for current deployments
Emphasise the possibility of expanding deployments to the 10-15m depth range to enhance survival

Reefs have the greatest probability of being clustered into Low if shallow, and High at
greater depths, with medium cluster reefs exhibiting a lower, and stable, variation in
assignment probability across depth ranges [@fig-ts-cluster-pdp, Panel B]. A large increase
in the probability of a reef being categorised as High can be seen between 10 and 12 meters
depth. This is at the edge of what is feasible with current technologies and restoration
approaches **\[TI: Citation here\]**, but indicates a depth range that could be adopted to
maximise longevity and benefits of, for example, coral seeding activities. With regard to
DHWs, assignment probability is roughly equal across the three categories up to 6 DHW. A
sharp delineation can be seen \> 6 DHW when comparing High with other categories, with lower
assignment probability (\~25%) compared to Medium and Low (\~40%) [@fig-ts-cluster-pdp,
Panel B].

@@

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced,
and exhibit overall similar DHW time series. While these reefs experience similar magnitude
DHWs based on SST, the proportion of the reefs' coral population impacted by heat stress
events depends on reef depth. Our results suggest that within shallow-water reef ecosystems
(non-mesophotic, depth \< 30m), depth contributes greatly to reef resilience and the ability
of the reef to maintain positive carbonate budgets under future conditions. While similar
findings have been demonstrated in the eastern Pacific [@smith2014], depth refugia analyses
have often focussed on the potential of mesophotic reefs (\> 30m deep) as thermal refugia
[@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019]
found no association between depth and the alleviation of heat stress across a range of
shallow reefs in the Pacific. Reefs at shallow depths experience no, or few, years with
coral cover exceeding levels likely required to maintain positive carbonate budgets.
Movement of shallow reefs towards negative carbonate budgets in the future may contribute to
phase shifts in reef dominant CaCO~3~ production. Shifting from coral driven to encrusting
algae driven production, or to bioerosion dominated systems with rapid erosion of physical
reef frameworks [@dudgeon2010; @perry2008]. While reefs at increased depths may be able to
maintain coral driven CaCO~3~ production and accumulation into the future for longer.

@@

Current reef restoration and intervention-investigation efforts on the GBR, such as
deployment of coral seeding and fragment devices, target reefs with suitable restoration
sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. One of the main
constraints on deployment depth is visual identification of target sites from the surface
for deployment and ongoing monitoring, in combination with factors such as deployment method
and tourism return value. Our results suggest that of the assessed reefs ocurring within the
1-10m depth range (1736 reefs), 62% (1082) of reefs fall into the low and medium clusters
(across all GCMs at the GBRMPA management region scale), with many unable to maintain
positive carbonate budgets under future conditions. Reef intervention planning is a complex
and challenging decision making process [@anthony2020], and our results suggest that
deployment site depth may be an important factor to additionally consider when assessing the
long-term resilience of deployment sites.

**\[TI note: Important to speak to interesting depth range at edge of where coral planting
might feasibly/viably occur with current approaches\]**

>>>>

Reefs have the greatest probability of being clustered into Low if shallow, and High at
greater depths, with medium cluster reefs exhibiting a lower, and stable, variation in
assignment probability across depth ranges [@fig-ts-cluster-pdp, Panel B]. A large increase
in the probability of a reef being categorised as High can be seen between 10 and 15 meters
depth. This is at the edge of the depth range used in current deployment efforts
[@howlett2021; @randall2021; @whitman2024], but indicates a depth range that could be
adopted to maximise longevity and benefits of, for example, coral seeding activities. With
regard to DHWs, assignment probability is roughly equal across the three categories up to 6
DHW. A sharp delineation can be seen \> 6 DHW when comparing High with other categories,
with lower assignment probability (\~25%) compared to Medium and Low (\~40%)
[@fig-ts-cluster-pdp, Panel B].

@@

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced,
and exhibit overall similar DHW time series. While these reefs experience similar magnitude
DHWs based on SST, the proportion of the reefs' coral population impacted by heat stress
events depends on reef depth. Reefs at deeper depths are likely to be clustered into High
cover clusters. Reefs at shallow depths experience no, or few, years with coral cover
exceeding levels likely required to maintain positive carbonate budgets. Movement of shallow
reefs towards negative carbonate budgets in the future may contribute to phase shifts in
reef dominant CaCO~3~ production. Shifting from coral driven to encrusting algae driven
production, or to bioerosion dominated systems with rapid erosion of physical reef
frameworks [@dudgeon2010; @perry2008]. While reefs at increased depths may be able to
maintain coral driven CaCO~3~ production and accumulation into the future for longer. Our
results suggest that within shallow-water reef ecosystems (non-mesophotic, depth \< 30m),
depth contributes greatly to reef resilience and the ability of the reef to maintain
positive carbonate budgets under future conditions. While similar findings have been
demonstrated in the Pacific [@smith2014; @baird2018] and Indo-Pacific [@bridge2014] regions,
depth refugia analyses have often focussed on the potential of mesophotic reefs (\> 30m
deep) as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas
et al. [-@venegas2019] found no association between depth and the alleviation of heat stress
across a range of shallow reefs in the Pacific.

@@

Current reef restoration and intervention-investigation efforts on the GBR, such as
deployment of coral seeding and fragment devices, target reefs with suitable restoration
sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are
no explicit justifications for this depth range, restoration depth is likely constrained by
snorkeling and diving logistical accessibility, visual identification of sites from the
surface, in combination with tourism return value. Our results suggest that of the assessed
reefs ocurring within the 1-10m depth range (1736 reefs), 62% (1082) of reefs fall into the
low and medium clusters (across all GCMs at the GBRMPA management region scale), with many
unable to maintain positive carbonate budgets under future conditions. Partial dependence
analysis revealed a range of depths from 10-15m where the probability of a reef being
clustered into the High cover category increased substatially. Considering this 10-15m depth
range when assessing possible restoration sites may enhance restoration success, while still
maintaining effective logistical access to sites. Reef intervention planning is a complex
and challenging decision making process [@anthony2020], and our results suggest that
deployment site depth may be an important factor to additionally consider when assessing the
long-term resilience of deployment sites.
